---
title: "Capstone Proposal"
subtitle: "Introduction"

author: 
  name: "M. Fazio"
  course: "Capstone Projects in Data Science"

date: today
date-format: "MMMM DD, YYYY"

format: pdf
    
bibliography: ../1-initiating/references.bib # file contains bibtex for references
citeproc: true

always_allow_html: true # this allows to get PDF with HTML features
self-contained: true

execute:
  cache: false
  freeze: false 
  warning: false
  message: false
  echo: false

editor: 
  markdown: 
    wrap: 72


---
```{r}
library(knitr)
library(reader)
library(tidyverse)

```

# Introduction

The eastern oyster (*Crassostrea virginica*) are viewed as a foundational species as they are directly responsible for creating and providing numerous ecosystem goods and services such as food and habitat production, shoreline stabilization and protection, water quality improvement, nutrient cycling, and carbon sequestration [@Geselbracht2024; @Grabowski2012; @Barnes2007]. Additionally, they provide numerous economic benefits such as supporting oyster fisheries, recreational harvesting, tourism, along with the associated employment that goes along with those  industries [@Geselbracht2024; @LaPeyre2021]. In one study, @Grabowski2012 estimated the average annual value to range from $10,325 to $99,421 per hectare (2011 Dollars).   

In recent decades, oyster reefs have experienced substantial declines, with only an estimated 15% remaining worldwide, and 50 to 80% being lost within the Gulf of (America) alone [@Pollack2012; @LaPeyre2021]. The decline is largely attributed to a combination of anthropogenic activities such us over harvesting, habitat alteration, sedimentation and water quality decline, disease and changing hydrology [@LaPeyre2021; @Geselbracht2024]. In response to these sharp declines, many restoration efforts have been undertaken with varying strategies and success rates; recently, planning documents released by the Deepwater Horizon Natural resource Damage Assessment Trustees called for large-scale development of reef networks to increase resilience [@Geselbracht2024]. 

**Oyster decline in the Pensacola Bay System**

On such system to experience oyster decline is the Pensacola Bay System (PBS). The PBS is a large estuarine system located in Escambia and Santa Rosa counties in northwest Florida. The system is comprised of multiple waterbody's including Pensacola Bay, Escambia Bay, Blackwater Bay, East Bay, and Santa Rosa Sound; and is influenced heavily by freshwater input from the Escambia, BLackwater and Yellow Rivers [@Christensen1997]. Historically, the PBS supported extensive oyster reefs that provided critical habitat for a variety of species and substantial economic value to the surrounding communities [@Christensen1997]. However, over the past several decades, the extent of oyster habitat within the PBS has declined significantly due to a combination of factors including over harvesting, habitat degradation, and water quality issues. With some estimates indicating a decline of 15,000ha to just 91.5 ha of oyster habitat between the years of 1883 and 2021 [@Geselbracht2024]. This decline has prompted efforts to restore oyster reefs within the PBS [@PPBEP_OysterRestoration; @TNC2021OysterRestoration], with multiple studies aiming to identify the most suitable areas within the estuary to focus restoration efforts [@Geselbracht2024; @Christensen1997; @PPBEP_OysterRestoration].

With limited resources available for restoration, it is critical to identify areas where chances of oyster survival and long-term reef establishment are at their highest, and where poor site selection could ultimately lead to unsatisfactory results. As with any ecosystem, the environmental conditions within the PBS vary greatly in time and space, and such variations can have great impacts on oyster survival and establishment. Therefore, it is important to understand the environmental factors that drive oyster occurrence so that scientist and policy makers can make informed decisions regarding said restoration site selection [@Pollack2012; @Barnes2007; @Geselbracht2024].

**Environmental controls on Eastern oyster habitat**

Oyster recruitment, survival, and reef establishment are influenced by complex interactions between numerous physical, chemical, and biological environmental factors. As such, determining the most influential factors is a multivariate problem that requires careful consideration of each factor and its interactions with others [@LaPeyre2021; @Pollack2012; @Barnes2007; @Christensen1997]. To help narrow the focus, the most commonly cited environmental factors were examined in the literature and are summarized here. 
   
Perhaps the most influential and widely reported environmental driver of oyster suitability is salinity. Representing an important environmental gradient within estuarine systems that influences physiology, recruitment, disease, predation, and survival. Other important characteristics often reported can be seen in @tbl-environmental-drivers and include: dissolved oxygen, temperature, depth, turbidity, and bottom type (substrate) [@LaPeyre2021; @Pollack2012; @Barnes2007; @Geselbracht2024; @Christensen1997]. 

::: {.landscape} 

```{r}
#| label: tbl-environmental-drivers
#| tbl-cap: "Common environmental drivers of Eastern Oyster habitat suitability."


oyster_drivers <- read_csv(
  "4-report/table_1_oyster_environmental_drivers.csv",
  show_col_types = FALSE
)

kable(
  oyster_drivers,
  format = "pipe",
  align = c("l", "l", "l", "c")
)
```

:::

It should be recognized that not all environmental factors are adequately spatially distributed or represented within the available datasets, and therefore careful consideration should be given when selecting the final dataset used for modeling habitat suitability [@LaPeyre2021; @Geselbracht2024; @Christensen1997].

**Environmentally Driven Habitat Modeling**

Quantifying the relationships between the distribution of species and the species-environment interactions have a long history in ecological research, having been described as far back as the late 17th early 18th century in which researchers observed biological patterns in terms of their relationships with geographical and/or environmental gradients [@Rushton2004; @Elith2009]. Since then, the propagation of large-scale spatial datasets, the development and integration of GIS, increased access to statistical packages and computational power have allowed for much more development in modeling complex species-environment interactions [@Johnson2005].  

**Habitat Suitability Indices (HSIs)**   

HSIs use expert-driven variable selection and classification methods that transform environmental characteristics into standardized estimates of habitat suitability, and have often been used as decision support tools in restoration frameworks due to their relatively low cost and flexibility [@Barnes2007; @Pollack2012; @Geselbracht2024]. 

Further, such models have already been used substantially in Eastern oyster restoration through out the Gulf Coast region in multiple different systems showing there applicability across spatial scales  [@Barnes2007; @Pollack2012; @Geselbracht2024; @Christensen1997]

Within the PBS, HSIs have been used to develop GIS-based and estuary-specific HSMs that integrate biological, physical, chemical, and management-related variables in order to identify suitable restoration areas with the system. In the mid 1990's @Christensen1997 developed a HSI for the eastern oyster in the PBS to demonstrate the model frameworks usefulness in illustrating species suitability in a system wide approach. @Geselbracht2024 then followed and further expanded on the approach incorporating management factors into the model, adding a layer of regulatory influence to the model output versus the purely environmental analysis of the prior. Additionally, the Pensacola and Perdido Bay Estuary Program recently built on-top of @Geselbracht2024 by incorporating some of their own data and refining some of the weighting [@PPBEP_OysterRestoration]. 

These previous studies alone highlight the strengths and weakness of the HSI approaches. Some of the strengths include interoperability, adaptability, and the integration of ecological knowledge, but the results entirely rely on expert decisions regarding variable selection, suitability functions, weighting and scoring, and how to aggregate the results. All of these present opportunity for differing outputs even if the same inputs were used by different groups [@Geselbracht2024; @Johnson2005].

**Species Distribution Modeling**

Species distribution models (SDMs) can be broadly defined as numerical tools that combine observations of species occurrence or abundance with environmental estimates [@Elith2009]. When combined with Geographic Information Systems (GIS), these predictive models have the ability to visualize the distribution and infer the suitability of species habitats. Further, these maps allow conservation professionals to concentrate and target their efforts in the hope to make more effective decisions [@Johnson2005]. 

**Species distribution models as an empirical alternative or complement**
   
   1. Introduce species distribution models (SDMs).
      - SDMs quantitatively relate known species distributions to environmental conditions.
      - They can be used for ecological explanation and spatial prediction.
      - Citation: [@Elith2009]
   
   2. Different habitat-modeling approaches can produce different spatial predictions.
      - Model selection and environmental-variable selection can influence areas identified as suitable.
      - Comparing modeling approaches can reveal areas of agreement and disagreement.
      - Citation: [@Johnson2005]
   
   3. Empirical SDMs provide an alternative method of estimating habitat relationships directly from observed species occurrence.
      - Establish the conceptual transition from HSI approaches to MaxEnt.
      - Citations: [@Elith2009; @Johnson2005]

6. **Maximum Entropy species distribution modeling**
   
   1. Introduce MaxEnt as a presence-background species distribution modeling framework.
      - Particularly useful where reliable absence observations are unavailable.
      - Relates environmental conditions at known occurrence locations to environmental conditions available throughout the modeling region.
      - Citations: [@Phillips2006; @Elith2011]
   
   2. Explain why MaxEnt is appropriate for Eastern oyster occurrence data in PBS.
      - FWC oyster-bed data provide known oyster occurrences without systematically sampled absence locations.
      - Presence-background modeling therefore fits the available data structure.
      - Citations: [@Phillips2006; @FWC2026Oysters]
   
   3. Explain important considerations when implementing MaxEnt.
      - Background extent and sampling.
      - Spatial sampling bias.
      - Predictor selection.
      - Regularization.
      - Feature classes and model complexity.
      - Citations: [@Phillips2009; @Merow2013; @Elith2011]
   
   4. Explain the need for model tuning and spatial validation.
      - MaxEnt should not be treated as an automated black-box model.
      - Model configuration should be evaluated systematically.
      - Spatially independent validation should be used where occurrence data exhibit spatial structure.
      - Citations: [@Kass2021; @Roberts2017; @Valavi2019]

7. **Research rationale, objectives, and hypotheses**
   
   1. Establish the primary research problem.
      - An estuary-specific HSI already exists for Eastern oyster restoration within PBS.
      - This study will independently develop an empirically fitted MaxEnt habitat-suitability model using observed oyster occurrences and environmental conditions.
      - Citations: [@Geselbracht2024; @Phillips2006]
   
   2. Primary objective.
      - Develop a spatially explicit MaxEnt SDM representing relative Eastern oyster habitat suitability throughout PBS.
   
   3. Secondary objective.
      - Quantitatively compare the MaxEnt suitability surface with the existing Pensacola Bay HSI.
   
   4. Supporting objectives.
      - Identify environmental predictors associated with observed oyster distribution.
      - Evaluate MaxEnt predictive performance using spatially separated validation data.
      - Identify geographic areas where MaxEnt and the HSI agree or disagree.
      - Evaluate whether MaxEnt provides a reproducible empirical complement to HSI-based restoration planning.
      - Citations: [@Johnson2005; @Geselbracht2024; @Merow2013]
   
   5. Research hypothesis.
      - The MaxEnt model is expected to identify patterns of habitat suitability consistent with established Eastern oyster habitat requirements.
      - Measurable agreement is expected between MaxEnt and the existing HSI, with spatial disagreement arising from differences between empirically fitted environmental relationships and predefined HSI suitability relationships.
      - Citations: [@Johnson2005; @Geselbracht2024]


# References