# Methods

```{r}
#| label: 2-load-packages-configs
#| echo: false

library(sf)
library(tidyverse)
library(tmap)
library(knitr)
library(tibble)

#load functions from source script
source("2-processing/scripts/process_wq_data.r")

# set tmap to static map
tmap_mode("plot")

```

## Data Processing and Analysis

### Software and Packages

All data processing, analysis, and modeling will be conducted using free
open source software and packages to maximize accessibility and
reproducibility. The majority of data manipulation and processing will
be performed in the R programming language using the tidyverse suite of
packages [@tidyverse; @r]. Any data operations that need to be
"spatially aware" will be performed using the Simple Features package
[@sf].

Analysis will be scripted and ran in Positron, an open-source integrated
development environment developed by Posit that supports R, Python,
scripts, notebooks, integrated consoles, along with interactive plots
and data viewers.

### Study Area

The study area for this project is the Pensacola Bay System which
includes portions of Pensacola Bay, Escambia Bay, East Bay and
Blackwater Bay. Boundaries of this study area were achieved by accessing
Florida Department of Protections (FDEP) Waterbod IDs feature layer
hosted on ArcGIS Online [@FDEP2005WBIDs]. ArcGIS allows direct access to
hosted feature services via several formats. To avoid any potential
compatibility issues, the GeoJSON format was chosen as this format is
non-proprietary. Once retrieved, the dataset was filtered to include
only relevant WBIDs (WBID = 548) and then combined to create a singular
study area polygon. Creating the study area polygon is an important
first step as it will be needed to extract relevant environmental data
from much larger datasets in the coming data gathering.

```{r}
#| label: 2-create-study-area
#| output: false

study_area <- get_study_area()

tm_shape(study_area) +
  tm_polygons(fill_alpha = 0, col = "red", lwd = 2) +
  tm_basemap("Esri.WorldStreetMap") +
  tm_layout(inner.margins = c(0.2, 0.2, 0.2, 0.2))

```

### Water Quality Portal Data

Water quality data is widely available for the entire United States via
the EPA's Water Quality Portal (WQP) which allows for a wide range of
data querying for any data as it relates to Water Quality Monitoring
programs (@wqp_2026). For this analysis, site data for Escambia and Santa Rosa County was queried using the WQP API then the study area polygon was used to extract only the relevent sites:

```{r}
#| echo: false
#| label: 2-fig-study-area
#| fig-cap: "Pensacola Bay System with monitoring sites"
#| fig-align: center

pbs_sites <- get_wq_sites(study_area)

## Add sites to map for visualization
tm_shape(pbs_sites) +
  tm_dots(
     fill = "red",
     size = 0.3
     ) +
tm_shape(study_area) +
  tm_polygons(
    fill_alpha = 0, 
    col = "black", 
    lwd = 2
    ) +
tm_basemap("Esri.WorldStreetMap") +
tm_layout(inner.margins = c(0.2, 0.2, 0.2, 0.2))
```

### Water Quality Data

Water quality data was selected based off of availability and relevance within literature. Each parameter was filtered to include just those measurements 
```{r}
#| label: 2-retrieve-process-wq-data

salinity <- get_wq_data(
  param = "Salinity",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "salinity",
  force = FALSE
  )

water_temp <- get_wq_data(
  param = "Temperature, water",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "water_temp",
  force = FALSE
  )

do_mgL <- get_wq_data(
  param = "Dissolved oxygen (DO)",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "do_mgL",
  force = FALSE
  )

chl_a_mgL <- get_wq_data(
  param = "Chlorophyll a, corrected for pheophytin",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "chl_a_mgL",
  force = FALSE
  )

turbidity <- get_wq_data(
  param = "Turbidity",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "turbidity",
  force = FALSE
  )  

```


```{r}
#| echo: false
#| label: 2-tbl-candidate-variables
#| tbl-cap: "Preliminary Environmental Variables Considered for the Eastern Oyster SDM"

data_dict <- read_csv("2-processing/data/data_dictionary.csv") 
  
kable(
  data_dict,
  align = c("l", "l", "l","l")
)
```


```{r}
#| label: 2-process-wq-data

sal <- proc_wq_data(salinity)
do <- proc_wq_data(do_mgL)
chla <- proc_wq_data(chl_a_mgL)
turb <- proc_wq_data(turbidity)
temp <- proc_wq_data(water_temp)
```


```{r}
#| label: 2-create-histogram-and-maps-to-view-distribution

wq_summary <- bind_rows(
  sal  %>% st_drop_geometry() %>% mutate(parameter = "Salinity (ppt)"),
  do   %>% st_drop_geometry() %>% mutate(parameter = "Dissolved Oxygen (mg/L)"),
  chla %>% st_drop_geometry() %>% mutate(parameter = "Chlorophyll-a (µg/L)"),
  turb %>% st_drop_geometry() %>% mutate(parameter = "Turbidity (FTU)"),
  temp %>% st_drop_geometry() %>% mutate(parameter = "Temperature (°C)")
)

ggplot(wq_summary, aes(x = mean)) +
  geom_histogram(
    bins = 20,
    color = "white",
    linewidth = 0.3
  ) +
  facet_wrap(
    ~ parameter,
    scales = "free",
    ncol = 2
  ) +
  labs(
    title = "Distribution of Warm-Season Water Quality Conditions",
    subtitle = "Site-level means for monitoring locations in the Pensacola Bay System",
    x = "Site Mean",
    y = "Number of Monitoring Locations"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# map points for each variable
ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = sal, aes(size = n, fill = mean),
  shape = 21,
    color = "white",
    stroke = 0.25
  ) +
  scale_fill_viridis_c(
    option = "viridis"
  ) +
  labs(
    title = "Warm-Season Salinity Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 

ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = temp, aes(size = n)) +
  labs(
    title = "Warm-Season Temperature Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 

ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = chla, aes(size = n)) +
  labs(
    title = "Warm-Season Chloraphyll-a Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 

ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = do, aes(size = n)) +
  labs(
    title = "Warm-Season Dissolved Oxygen Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 

ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = turb, aes(size = n)) +
  labs(
    title = "Warm-Season Turbidity Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 
```


::: {.landscape}
```{r}
#|label: 2-summary-table

wq_summary_table <- wq_summary %>%
  group_by(parameter) %>%
  summarise(
    stations = n(),
    observations = sum(n),
    overall_mean = mean(mean, na.rm = TRUE),
    overall_sd = sd(mean, na.rm = TRUE),
    min = min(mean, na.rm = TRUE),
    q1 = quantile(mean, 0.25, na.rm = TRUE),
    median = median(mean, na.rm = TRUE),
    q3 = quantile(mean, 0.75, na.rm = TRUE),
    max = max(mean, na.rm = TRUE),
    first_date = min(first_date, na.rm = TRUE),
    last_date = max(last_date, na.rm = TRUE),
    .groups = "drop"
  )
  

kable(
  wq_summary_table,
  caption = "Summary Statistics for Warm-Season Water Quality Data",
  col.names = c(
    "Parameter",
    "Stations",
    "Observations",
    "Mean",
    "SD",
    "Min",
    "Q1",
    "Median",
    "Q3",
    "Max",
    "First Sample",
    "Last Sample"
  ),
  align = c("l", "c", "c", "c"),
  digits = 2,
  booktabs = TRUE
) 
```

:::

## Spatial interpolation

Points are interpolated into prediction rasters using inverse path distance weighting  (IPDW) (package idpw). Blah blah blah blah

