# Methods

```{r}
#| label: 2-load-packages-configs
#| echo: false

library(sf)
library(tidyverse)
library(tmap)
library(knitr)
library(tibble)

#load functions from source script
source("3-scripts/process_wq_data.r")

# set tmap to static map
tmap_mode("plot")

```

## Data Processing and Analysis

### Software and Packages

All data processing, analysis, and modeling will be conducted using free
open source software and packages to maximize accessibility and
reproducibility. The majority of data manipulation and processing will
be performed in the R programming language using the tidyverse suite of
packages [@tidyverse; @r]. Any data operations that need to be
"spatially aware" will be performed using the Simple Features package
[@sf].

Analysis will be scripted and ran in Positron, an open-source integrated
development environment developed by Posit that supports R, Python,
scripts, notebooks, integrated consoles, along with interactive plots
and data viewers.

### Study Area

The study area for this project is the Pensacola Bay System which
includes portions of Pensacola Bay, Escambia Bay, East Bay and
Blackwater Bay. Boundaries of this study area were achieved by accessing
Florida Department of Protections (FDEP) Waterbod IDs feature layer
hosted on ArcGIS Online [@FDEP2005WBIDs]. ArcGIS allows direct access to
hosted feature services via several formats. To avoid any potential
compatibility issues, the GeoJSON format was chosen as this format is
non-proprietary. Once retrieved, the dataset was filtered to include
only relevant WBIDs (WBID = 548) and then combined to create a singular
study area polygon. Creating the study area polygon is an important
first step as it will be needed to extract relevant environmental data
from much larger datasets in the coming data gathering.

```{r}
#| label: 2-create-study-area
#| output: false

study_area <- get_study_area()

tm_shape(study_area) +
  tm_polygons(fill_alpha = 0, col = "red", lwd = 2) +
  tm_basemap("Esri.WorldStreetMap") +
  tm_layout(inner.margins = c(0.2, 0.2, 0.2, 0.2))

```

### Water Quality Portal Data

Water quality data is widely available for the entire United States via
the EPA's Water Quality Portal (WQP) which allows for a wide range of
data querying for any data as it relates to Water Quality Monitoring
programs (@wqp_2026). For this analysis, site data for Escambia and Santa Rosa County was queried using the WQP API then the study area polygon was used to extract only the relevent sites:

```{r}
#| echo: false
#| label: 2-fig-study-area
#| fig-cap: "Pensacola Bay System with monitoring sites"
#| fig-align: center

pbs_sites <- get_wq_sites(study_area)

## Add sites to map for visualization
tm_shape(pbs_sites) +
  tm_dots(
     fill = "red",
     size = 0.3
     ) +
tm_shape(study_area) +
  tm_polygons(
    fill_alpha = 0, 
    col = "black", 
    lwd = 2
    ) +
tm_basemap("Esri.WorldStreetMap") +
tm_layout(inner.margins = c(0.2, 0.2, 0.2, 0.2))
```

### Water Quality Data

Water quality data was selected based off of availability and relevance within literature. Each parameter was filtered to include just those measurements 
```{r}
#| label: 2-retrieve-process-wq-data

salinity <- get_wq_data(
  param = "Salinity",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "salinity",
  force = FALSE
  )

water_temp <- get_wq_data(
  param = "Temperature, water",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "water_temp",
  force = FALSE
  )

do_mgL <- get_wq_data(
  param = "Dissolved oxygen (DO)",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "do_mgL",
  force = FALSE
  )

chl_a_mgL <- get_wq_data(
  param = "Chlorophyll a, corrected for pheophytin",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "chl_a_mgL",
  force = FALSE
  )

turbidity <- get_wq_data(
  param = "Turbidity",
  from_date = "2005-01-01",
  to_date = "2025-01-01",
  sites = pbs_sites,
  output_name = "turbidity",
  force = FALSE
  )  

```


```{r}
#| echo: false
#| label: 2-tbl-candidate-variables
#| tbl-cap: "Preliminary Environmental Variables Considered for the Eastern Oyster SDM"

data_dict <- read_csv("2-data/data_dictionary.csv") 
  
kable(
  data_dict,
  align = c("l", "l", "l","l")
)
```


```{r}
#| label: 2-process-wq-data

sal <- proc_wq_data(salinity)
do <- proc_wq_data(do_mgL)
chla <- proc_wq_data(chl_a_mgL)
turb <- proc_wq_data(turbidity)
temp <- proc_wq_data(water_temp)
```


```{r}
#| label: 2-create-histogram-and-maps-to-view-distribution

wq_summary <- bind_rows(
  sal  %>% st_drop_geometry() %>% mutate(parameter = "Salinity (ppt)"),
  do   %>% st_drop_geometry() %>% mutate(parameter = "Dissolved Oxygen (mg/L)"),
  chla %>% st_drop_geometry() %>% mutate(parameter = "Chlorophyll-a (µg/L)"),
  turb %>% st_drop_geometry() %>% mutate(parameter = "Turbidity (FTU)"),
  temp %>% st_drop_geometry() %>% mutate(parameter = "Temperature (°C)")
)

ggplot(wq_summary, aes(x = mean)) +
  geom_histogram(
    bins = 20,
    color = "white",
    linewidth = 0.3
  ) +
  facet_wrap(
    ~ parameter,
    scales = "free",
    ncol = 2
  ) +
  labs(
    title = "Distribution of Warm-Season Water Quality Conditions",
    subtitle = "Site-level means for monitoring locations in the Pensacola Bay System",
    x = "Site Mean",
    y = "Number of Monitoring Locations"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# map points for each variable
ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = sal, aes(size = n, fill = mean),
  shape = 21,
    color = "white",
    stroke = 0.25
  ) +
  scale_fill_viridis_c(
    option = "viridis"
  ) +
  labs(
    title = "Warm-Season Salinity Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 

ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = temp, aes(size = n)) +
  labs(
    title = "Warm-Season Temperature Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 

ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = chla, aes(size = n)) +
  labs(
    title = "Warm-Season Chloraphyll-a Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 

ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = do, aes(size = n)) +
  labs(
    title = "Warm-Season Dissolved Oxygen Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 

ggplot() +
  geom_sf(data = study_area, fill = "grey95") +
  geom_sf(data = turb, aes(size = n)) +
  labs(
    title = "Warm-Season Turbidity Monitoring Locations",
    subtitle = "Point size represents number of observations"
  ) +
  theme_minimal() 
```


::: {.landscape}
```{r}
#|label: 2-summary-table

wq_summary_table <- wq_summary %>%
  group_by(parameter) %>%
  summarise(
    stations = n(),
    observations = sum(n),
    overall_mean = mean(mean, na.rm = TRUE),
    overall_sd = sd(mean, na.rm = TRUE),
    min = min(mean, na.rm = TRUE),
    q1 = quantile(mean, 0.25, na.rm = TRUE),
    median = median(mean, na.rm = TRUE),
    q3 = quantile(mean, 0.75, na.rm = TRUE),
    max = max(mean, na.rm = TRUE),
    first_date = min(first_date, na.rm = TRUE),
    last_date = max(last_date, na.rm = TRUE),
    .groups = "drop"
  )
  

kable(
  wq_summary_table,
  caption = "Summary Statistics for Warm-Season Water Quality Data",
  col.names = c(
    "Parameter",
    "Stations",
    "Observations",
    "Mean",
    "SD",
    "Min",
    "Q1",
    "Median",
    "Q3",
    "Max",
    "First Sample",
    "Last Sample"
  ),
  align = c("l", "c", "c", "c"),
  digits = 2,
  booktabs = TRUE
) 
```

:::

## Spatial interpolation

Points are interpolated into prediction rasters using inverse path distance weighting  (IPDW) (package idpw). Blah blah blah blah

# Methods 

**Study design and team status**

This capstone is being completed as an individual project, an all data acquisition, processing, analysis and interpolation will be completed by the author. 


   2. Describe the overall analytical workflow.
      - Environmental and occurrence data acquisition.
      - Data QA/QC and preprocessing.
      - Creation of continuous environmental predictor surfaces.
      - Oyster occurrence preparation.
      - Background sampling.
      - Predictor screening.
      - MaxEnt tuning and fitting.
      - Spatial model validation.
      - HSI comparison.
      - Citations: [@Phillips2006; @Merow2013; @Kass2021]

2. **Study area**
   
   1. Define the Pensacola Bay System modeling extent.
      - Delineate the study area using selected Florida Department of Environmental Protection Waterbody Identification (WBID) polygons.
      - Describe the primary embayments included within the analysis.
      - Citations: [@FDEP2005WBIDs; @Geselbracht2024]
   
   2. Restrict the modeling domain to aquatic habitat.
      - Background and prediction locations should represent environments potentially available to Eastern oysters.
      - Terrestrial areas should be excluded from the modeling domain.
      - Citations: [@Phillips2009; @Merow2013]
   
   3. Maintain spatial compatibility with the existing Pensacola Bay HSI where possible.
      - A comparable geographic extent will support subsequent spatial comparison.
      - Citation: [@Geselbracht2024]
   
   4. Develop a study-area map.
      - PBS boundary.
      - Major embayments.
      - Oyster occurrence locations.
      - Environmental monitoring locations.

3. **Data acquisition and dataset inventory**
   
   1. Eastern oyster occurrence data.
      - Source data from the FWC-FWRI Oyster Beds in Florida dataset.
      - Record dataset version and acquisition date.
      - Document original geometry type.
      - Document number of source features and number retained within the study area.
      - Document temporal/source-data information where available.
      - Citation: [@FWC2026Oysters]
   
   2. Water-quality observations.
      - Retrieve observations from the Water Quality Portal.
      - Use the `dataRetrieval` package and associated WQP services.
      - Restrict queries to relevant counties, estuarine monitoring locations, selected parameters, and the study period.
      - Citations: [@wqp_2026; @dataRetrieval2026]
   
   3. Document water-quality dataset size and temporal coverage.
      - Dissolved oxygen: approximately 145 stations and 3,072 observations.
      - Salinity: approximately 135 stations and 3,401 observations.
      - Temperature: approximately 145 stations and 3,442 observations.
      - Turbidity: approximately 122 stations and 1,563 observations.
      - Chlorophyll-a: approximately 48 stations and 754 observations.
      - Current temporal coverage is approximately 2005--2024.
      - Citations for source data: [@wqp_2026; @dataRetrieval2026]
   
   4. Evaluate candidate predictors based on spatial coverage.
      - Chlorophyll-a is currently expected to be excluded from interpolation because available stations do not provide adequate system-wide spatial coverage.
      - Salinity, dissolved oxygen, temperature, and turbidity currently provide substantially stronger monitoring coverage.
      - Ecological rationale for candidate variables: [@LaPeyre2021; @Geselbracht2024]
   
   5. Bathymetry.
      - Obtain Pensacola Bay estuarine bathymetry from NOAA.
      - Document raster resolution, vertical datum, geographic extent, and preprocessing requirements.
      - Citation: [@NOAA_2018]
   
   6. Substrate or bottom type.
      - Identify and acquire a suitable spatial dataset describing bottom characteristics.
      - Document spatial resolution, classes, temporal source, and coverage.
      - Ecological justification: [@Barnes2007; @Geselbracht2024]
      - **Citation needed for final substrate dataset.**
   
   7. Existing Pensacola Bay HSI.
      - Obtain the original GIS/raster output used by Geselbracht et al.
      - Document raster score range, projection, extent, resolution, and NoData handling.
      - Citation for methodology: [@Geselbracht2024]
      - **Add a separate dataset citation once the original HSI GIS product is acquired.**

4. **Water-quality preprocessing**
   
   1. Standardize measurement characteristics and units.
      - Avoid combining fundamentally different measurement types that share similar characteristic names.
      - Retain comparable units and analytical forms for each predictor.
      - Citations: [@wqp_2026; @dataRetrieval2026]
   
   2. Restrict observations to warm-season conditions.
      - Current working definition: May through October.
      - Use seasonal filtering to characterize environmental conditions relevant to warm-season oyster stress and habitat suitability.
      - Citation for Pensacola seasonal precedent: [@Geselbracht2024]
   
   3. Generate station-level environmental summaries.
      - Calculate warm-season mean.
      - Calculate standard deviation.
      - Record total observation count.
      - Record number of unique years represented.
      - Record first and last sample date.
      - Consider calculating site-year warm-season means prior to calculating long-term site means to avoid over-weighting heavily sampled individual years.
   
   4. Perform QA/QC.
      - Identify potential unit inconsistencies.
      - Investigate implausible or extreme values.
      - Document removal or correction criteria.
      - Retain original raw data separately from processed datasets.
   
   5. Convert monitoring data to spatial features.
      - Use `sf` for spatial-vector processing.
      - Project observations to NAD 1983 (2011) StatePlane Florida North FIPS 0903 in meters.
      - Citation: [@sf]

5. **Generation of continuous environmental predictor surfaces**
   
   1. Assess monitoring-station spatial coverage prior to interpolation.
      - Map station distribution for each parameter.
      - Identify spatial clustering.
      - Identify portions of PBS lacking observations.
      - Exclude variables where prediction would consist primarily of spatial extrapolation.
   
   2. Account for coastal barriers and complex estuarine geometry.
      - Straight-line Euclidean distance may incorrectly represent spatial relationships between monitoring locations separated by peninsulas, islands, or narrow water connections.
      - Barrier-aware interpolation will be evaluated where feasible.
      - **Suggested citation: Stachelek and Madden (2015).**
   
   3. Evaluate inverse path distance weighting (IPDW) as a candidate interpolation method.
      - IPDW uses aquatic path distance rather than straight-line Euclidean distance.
      - This approach is specifically suited to coastal water-quality interpolation where land barriers affect connectivity.
      - **Suggested citation: Stachelek and Madden (2015).**
   
   4. Compare candidate interpolation approaches.
      - Potential methods include:
        - Inverse path distance weighting.
        - Conventional inverse distance weighting.
        - Ordinary kriging.
      - Use spatially withheld monitoring locations to evaluate predictive performance.
      - Compare RMSE, MAE, and bias.
      - Select the final interpolation method based on predictive performance and ecological appropriateness.
      - Citations: [@Roberts2017; @Valavi2019]
   
   5. Produce final environmental predictor rasters.
      - Use a common projected coordinate system.
      - Use a common cell size.
      - Use common raster origin and alignment.
      - Mask predictions to aquatic habitat.
      - Document final raster resolution.
      - Citation for Pensacola spatial-modeling precedent: [@Geselbracht2024]

6. **Oyster occurrence preparation**
   
   1. Clip FWC oyster-bed features to the study area.
      - Remove occurrences outside the PBS modeling extent.
      - Citation: [@FWC2026Oysters]
   
   2. Convert mapped oyster-bed polygons into representative MaxEnt presence observations.
      - Use a reproducible procedure to avoid overrepresenting large or heavily mapped oyster polygons.
      - Document the initial number of oyster polygons and final number of presence observations.
      - Citations: [@Phillips2009; @Merow2013]
   
   3. Evaluate spatial clustering of presence observations.
      - Apply spatial thinning where necessary.
      - Thinning should be based on sampling bias and spatial dependence rather than arbitrary data reduction.
      - Citations: [@Phillips2009; @Merow2013]

7. **Background sampling**
   
   1. Define the accessible/background modeling region.
      - Restrict background sampling to aquatic habitat within PBS.
      - Require complete environmental predictor coverage.
      - Exclude terrestrial areas and locations outside the modeled environmental domain.
      - Citations: [@Phillips2009; @Merow2013]
   
   2. Generate MaxEnt background locations.
      - Document the number of background points.
      - Sample from the defined accessible region rather than an arbitrary rectangular extent.
      - Citations: [@Phillips2009; @Merow2013]
   
   3. Address occurrence-sampling bias.
      - Evaluate whether spatial thinning, bias weighting, or restricted background sampling is necessary.
      - Citation: [@Phillips2009]

8. **Predictor screening**
   
   1. Assemble final candidate predictor stack.
      - Salinity.
      - Dissolved oxygen.
      - Temperature.
      - Turbidity.
      - Bathymetry/depth.
      - Substrate/bottom type, if suitable data are obtained.
      - Citations: [@LaPeyre2021; @Barnes2007; @Pollack2012; @Geselbracht2024]
   
   2. Evaluate correlation and redundancy among predictors.
      - Calculate pairwise correlations.
      - Identify highly correlated environmental variables.
      - Retain predictors based on ecological relevance, data quality, and interpretability.
      - **Suggested citation: Dormann et al. (2013).**

9. **MaxEnt model fitting and tuning**
   
   1. Fit MaxEnt using occurrence locations, background samples, and the finalized environmental predictor stack.
      - Citations: [@Phillips2006; @Elith2011]
   
   2. Tune model complexity rather than relying exclusively on default settings.
      - Evaluate multiple regularization multipliers.
      - Evaluate multiple feature-class combinations.
      - Citations: [@Merow2013; @Kass2021]
   
   3. Use the `ENMeval` framework for model tuning and evaluation.
      - Citation: [@Kass2021]
   
   4. Select the final MaxEnt configuration.
      - Consider predictive performance.
      - Consider omission rates.
      - Consider model complexity.
      - Citations: [@Kass2021; @Merow2013]
   
   5. Generate the final continuous habitat-suitability surface.
      - Retain continuous predictions for primary comparison with the HSI.
      - Citations: [@Elith2011; @Merow2013]
   
   6. Evaluate environmental-variable influence.
      - Examine permutation importance.
      - Examine response curves.
      - Evaluate other supported measures of predictor contribution.
      - Interpret associations as predictive relationships rather than direct causal effects.
      - Citations: [@Elith2011; @Merow2013]

10. **Spatial model validation**
    
    1. Use spatial rather than purely random cross-validation.
       - Nearby oyster occurrences may experience highly similar environmental conditions.
       - Random partitions may therefore produce overly optimistic estimates of predictive performance.
       - Citations: [@Roberts2017; @Valavi2019]
    
    2. Divide oyster occurrences into spatially separated validation folds.
       - Select spatial-block dimensions based on observed spatial structure.
       - Citation: [@Valavi2019]
    
    3. Evaluate models against withheld spatial folds.
       - Report appropriate presence-background performance metrics.
       - Consider omission rate and AUC where appropriate.
       - Report variation in model performance among folds.
       - Citations: [@Phillips2006; @Kass2021]

11. **Comparison with the existing Habitat Suitability Index**
    
    1. Harmonize the MaxEnt and HSI raster surfaces.
       - Match geographic extent.
       - Match coordinate reference system.
       - Match cell size and alignment.
       - Ensure suitability values are directly comparable.
       - Citations: [@Geselbracht2024; @Johnson2005]
    
    2. Compare continuous habitat-suitability predictions.
       - Calculate cell-by-cell correlation.
       - Consider Pearson and/or Spearman correlation depending on model-output distributions.
       - Citation: [@Johnson2005]
    
    3. Generate a spatial difference surface.
       
       $$
       D_i = S_{MaxEnt,i} - S_{HSI,i}
       $$
       
       - Values near zero represent agreement.
       - Positive values identify locations where MaxEnt predicts relatively greater suitability.
       - Negative values identify locations where the HSI predicts relatively greater suitability.
       - Citation: [@Johnson2005]
    
    4. Evaluate categorical agreement if suitability classes are created.
       - Both models predict high suitability.
       - Both models predict low suitability.
       - MaxEnt predicts high suitability while HSI predicts low suitability.
       - HSI predicts high suitability while MaxEnt predicts low suitability.
       - Consider an agreement statistic if appropriate.
       - Citation: [@Johnson2005]
    
    5. Interpret disagreement between models.
       - Determine whether disagreement is geographically concentrated.
       - Evaluate whether disagreement corresponds with particular environmental predictors.
       - Consider differences in data availability and modeling assumptions.
       - Avoid treating either model as automatically representing ground truth.
       - Citations: [@Johnson2005; @Geselbracht2024]

12. **Software and reproducibility**
    
    1. Conduct data processing, statistical analysis, and modeling in R.
       - Citation: [@r]
    
    2. Use scripted workflows for data transformation and spatial processing.
       - `tidyverse` for data manipulation.
       - `sf` for vector spatial processing.
       - Citations: [@tidyverse; @sf]
    
    3. Maintain reproducibility throughout the workflow.
       - Record package versions.
       - Set random seeds where applicable.
       - Record data-acquisition dates.
       - Document model settings and tuning parameters.
       - Preserve raw data separately from processed data.
       - Citations for dynamic data sources: [@dataRetrieval2026; @wqp_2026; @FWC2026Oysters]